E-Graphs With Bindings

### 2025-01-27 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
clean-up

@@

Another distinct challenge is presented by $\beta$-reduction, which is handled using \emph{explicit substitutions}. 
The implementations of $\beta$-reduction in e-graphs do not go into much detail regarding explicit substitution, and in particular do not include rules like $f[x/1][y/2] = f[y/2][x/1]$ for non-interfering substitutions, while such rules are absorbed by graph equivalence~\cite{accattoli2014nonstandard}.

>>>>

% change red to orange
\definecolor{e-color}{named}{orange}
\definecolor{closed-color}{named}{blue}

@@

Another distinct challenge is presented by $\beta$-reduction, which is handled using \emph{explicit substitutions}. 
The implementations of $\beta$-reduction in e-graphs do not go into much detail regarding explicit substitution, and in particular do not include rules like $[x\!\mapsto\!1][y\!\mapsto\!2]f\!=\![y\!\mapsto\!2][x\!\mapsto\!1]f$ for non-interfering substitutions, while such rules are absorbed by graph equivalence~\cite{accattoli2014nonstandard}.

### 2025-01-27 LICS_2025/e_graphs_with_bindings/sections/appendix.tex
clean-up

In constructing the pushout the functional versions of e-hypergraph relations will be useful.
\begin{remark}
    $\textcolor{red}{<}$, $\textcolor{blue}{<}$ and $\consistency$ can be considered as (partial) functions defined on $V_{\mathcal{F}} + E_{\mathcal{F}}$, \textit{i.e.}, on the coproduct of vertices and edges.
    To make things well-typed, we will use corresponding coproduct injections $\iota_{V} : {V_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ and $\iota_{E} : {E_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ when passing either a vertex or an edge into these functions.
    For example, an immediate successor of a vertex $x$ can be written functionally as $\textcolor{red}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))$.
\end{remark}

@@

            \item When $x$ is not a top-level vertex such that $\textcolor{red}{<}(\iota_{V}(x))$ is defined, 
                \[
                \phi_{E}(\textcolor{red}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{red}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(\textcolor{red}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{red}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.
                \item When $x$ is not a top-level vertex such that $\textcolor{blue}{<}(\iota_{V}(x))$ is defined, 
                \[
                \phi_{E}(\textcolor{blue}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{blue}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(\textcolor{blue}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{blue}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.
                \item
            When $x \in E_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
            \]
            where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}] : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
            \item When $x \in V_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x)).
            \]
            \end{enumerate}
\end{definition}

@@

We next explicitly construct a pushout.
Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{red}{<_{X+Y}}, \textcolor{blue}{<_{X+Y}}\}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{red}{<_{X+Y}}, \textcolor{blue}{<_{X+Y}}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\\
          &\}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

\begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = (&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &\textcolor{red}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
            &\textcolor{blue}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        &)    
\end{align*}

@@

Recall that $\textcolor{red}{<}$ is essentially a transitive closure of $\textcolor{red}{<^{\mu}}$ and hence we can first define 
    \[\textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e])
    \] and 
    \[
        \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u])
    \] where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    We will consider several cases.
    First, assume that $u$ is an element of $V_{X+Y}$ and
    \begin{enumerate}
        \item If there exists $v$ such that $\textcolor{red}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v))$ is defined and $u \sim v$
              we let
              \[
                \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u])) = [\textcolor{red}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v))]
              \]
        \item \label{def:child_respects_connectivity} $u$ has no pre-image in $V_{Z}$ and $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined (\textit{i.e.}, $[u) = \varnothing$).
              If there exists $v$ such that $\textcolor{red}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an \textit{undirected} path from $[u]$ to $[v]$, then we define
              \[ 
                \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
              \]
    \end{enumerate}
    % Note that the existence of a path in the definition (2) implies that there exists $v' \sim v$ such that there is a path from $u$ to $v'$ and furthermore $v \not = v'$.
    % If $v' = v$ then it would mean that $[u) \not = \varnothing$ since $[v) \not = \varnothing$ and there is a path from $u$ to $v$.
    % There is no case when $u$ has a pre-image in $V_{Z}$ and yet $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e$ and there is a path from $[u]$ to $[v]$.
    % The existence of a path would mean that there is $u' \sim u$ and $v' \sim v$ such that there is a path from $u'$ to $v'$ and since $u$ has a pre-image in $V_{Z}$, $u = f_{V};\iota_{1,V}(z_1)$.
    % $v' \sim v$ means that both $v'$ and $v$ have a pre-image in $V_{Z}$ (unless $v' = v$ in which case $[u') \not = \varnothing$ as there is a path from $u'$ to $v'$) and since $u = f_{V};\iota_{1,V}(z_1)$ and $[u) = \varnothing$ $f_{V};\iota_{1,V}(z) = \varnothing$ for all $z$.
    % Since $[v) \not = \varnothing$ and $v' \sim v$, $v$ is necessarily in the image of $g_{V};\iota_{2,V}$ and for all $z$ $[g_{V};\iota_{2,V}(z)) \not = \varnothing$ as well as for some $z'$ such that $u'' = g_{V};\iota_{2,V}(z')$ and $u'' \sim_{S} u$ which would mean that the case (1) is applicable to $u$.
    % Similarly, when $u = g_{V};\iota_{2,V}(z)$.

    Otherwise, we leave $\textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))$ undefined.
    % \question{In the case when images of $f$ and $g$ are top-level the function is undefined. Shall we introduce bottom?}
    Now, assume that $e$ is an element of $E_{X+Y}$.
    Consider two cases.
    \begin{enumerate}
    \item  $\textcolor{red}{<_{X+Y}^{\mu}}(\iota_{E_{X} + E_{Y}}(e)) = e'$.
            Then we define
        \[
            \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = [\textcolor{red}{<_{X+Y}^{\mu}}(\iota_{E_{X} + E_{Y}}(e))]
        \]
    \item $[e) = \varnothing$ and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an undirected path from $[e]$ to $[v]$, we define 
        \[
            \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = \textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
        \] 
    \end{enumerate}
    Otherwise we leave $\textcolor{red}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{E_{X} + E_{Y}}{\sim}([e]))$ undefined.
    % We will also build this relation in steps, first we do steps for vertices (1) - (3), then step (1) for edges and finally, we will interleave step (2) for edges and step (4) for vertices until fixed point.
    % Essentially, this interleaving means that vertices inherit the relation from incident edges and vice versa.
    Clearly, all the cases above are disjoint.
    We similarly define $\textcolor{blue}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.

@@

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{blue}{<_{\sfrac{X+Y}{\sim}}}^{\mu}$ that is constructed analogously to $\textcolor{red}{<_{\sfrac{X+Y}{\sim}}}^{\mu}$.
\end{proof}
The construction can be illustrated with the example in Figure~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5.5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-2.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-1.5, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
such that $\textcolor{blue}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{blue}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{blue}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Figure~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

@@

Consider an example of DPOI rewriting in Figure~\ref{fig:dpoi-example} as defined in Definition~\ref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of Definition~\ref{def:dpoi-e}.

>>>>

In constructing the pushout the functional versions of e-hypergraph relations will be useful.
\begin{remark}
    $\textcolor{e-color}{<}$, $\textcolor{closed-color}{<}$ and $\consistency$ can be considered as (partial) functions defined on $V_{\mathcal{F}} + E_{\mathcal{F}}$, \textit{i.e.}, on the coproduct of vertices and edges.
    To make things well-typed, we will use corresponding coproduct injections $\iota_{V} : {V_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ and $\iota_{E} : {E_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ when passing either a vertex or an edge into these functions.
    For example, an immediate successor of a vertex $x$ can be written functionally as $\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))$.
\end{remark}

@@

            \item When $x$ is not a top-level vertex such that $\textcolor{e-color}{<}(\iota_{V}(x))$ is defined, 
                \[
                \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.
                \item When $x$ is not a top-level vertex such that $\textcolor{closed-color}{<}(\iota_{V}(x))$ is defined, 
                \[
                \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
                \]
                and
                \[
                \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))  
                \] when $x$ is a not top-level edge.
                \item
            When $x \in E_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
            \]
            where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}] : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
            \item When $x \in V_{\mathcal{F}}$
            \[
                [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x)).
            \]
            \end{enumerate}
\end{definition}

@@

We next explicitly construct a pushout.
Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\\
          &\}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

\begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = (&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &\textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
            &\textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        &)    
\end{align*}

@@

Recall that $\textcolor{e-color}{<}$ is essentially a transitive closure of $\textcolor{e-color}{<^{\mu}}$ and hence we can first define 
    \[\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e])
    \] and 
    \[
        \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u])
    \] where $\iota_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    We will consider several cases.
    First, assume that $u$ is an element of $V_{X+Y}$ and
    \begin{enumerate}
        \item If there exists $v$ such that $\textcolor{e-color}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v))$ is defined and $u \sim v$
              we let
              \[
                \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u])) = [\textcolor{e-color}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v))]
              \]
        \item \label{def:child_respects_connectivity} $u$ has no pre-image in $V_{Z}$ and $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined (\textit{i.e.}, $[u) = \varnothing$).
              If there exists $v$ such that $\textcolor{e-color}{<_{X+Y}^{\mu}}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an \textit{undirected} path from $[u]$ to $[v]$, then we define
              \[ 
                \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
              \]
    \end{enumerate}
    % Note that the existence of a path in the definition (2) implies that there exists $v' \sim v$ such that there is a path from $u$ to $v'$ and furthermore $v \not = v'$.
    % If $v' = v$ then it would mean that $[u) \not = \varnothing$ since $[v) \not = \varnothing$ and there is a path from $u$ to $v$.
    % There is no case when $u$ has a pre-image in $V_{Z}$ and yet $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(u))$ is undefined and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e$ and there is a path from $[u]$ to $[v]$.
    % The existence of a path would mean that there is $u' \sim u$ and $v' \sim v$ such that there is a path from $u'$ to $v'$ and since $u$ has a pre-image in $V_{Z}$, $u = f_{V};\iota_{1,V}(z_1)$.
    % $v' \sim v$ means that both $v'$ and $v$ have a pre-image in $V_{Z}$ (unless $v' = v$ in which case $[u') \not = \varnothing$ as there is a path from $u'$ to $v'$) and since $u = f_{V};\iota_{1,V}(z_1)$ and $[u) = \varnothing$ $f_{V};\iota_{1,V}(z) = \varnothing$ for all $z$.
    % Since $[v) \not = \varnothing$ and $v' \sim v$, $v$ is necessarily in the image of $g_{V};\iota_{2,V}$ and for all $z$ $[g_{V};\iota_{2,V}(z)) \not = \varnothing$ as well as for some $z'$ such that $u'' = g_{V};\iota_{2,V}(z')$ and $u'' \sim_{S} u$ which would mean that the case (1) is applicable to $u$.
    % Similarly, when $u = g_{V};\iota_{2,V}(z)$.

    Otherwise, we leave $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))$ undefined.
    % \question{In the case when images of $f$ and $g$ are top-level the function is undefined. Shall we introduce bottom?}
    Now, assume that $e$ is an element of $E_{X+Y}$.
    Consider two cases.
    \begin{enumerate}
    \item  $\textcolor{e-color}{<_{X+Y}^{\mu}}(\iota_{E_{X} + E_{Y}}(e)) = e'$.
            Then we define
        \[
            \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = [\textcolor{e-color}{<_{X+Y}^{\mu}}(\iota_{E_{X} + E_{Y}}(e))]
        \]
    \item $[e) = \varnothing$ and there exists $v$ such that $<_{X+Y}^{\mu}(\iota_{V_{X} + V_{Y}}(v)) = e'$ and such that there is an undirected path from $[e]$ to $[v]$, we define 
        \[
            \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[e]) = \textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v])
        \] 
    \end{enumerate}
    Otherwise we leave $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}(\iota_{E_{X} + E_{Y}}{\sim}([e]))$ undefined.
    % We will also build this relation in steps, first we do steps for vertices (1) - (3), then step (1) for edges and finally, we will interleave step (2) for edges and step (4) for vertices until fixed point.
    % Essentially, this interleaving means that vertices inherit the relation from incident edges and vice versa.
    Clearly, all the cases above are disjoint.
    We similarly define $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.

@@

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5.5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-2.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-1.5, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Fig.~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

@@

Consider an example of DPOI rewriting in Fig.~\ref{fig:dpoi-example} as defined in Definition~\ref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of Definition~\ref{def:dpoi-e}.
